---
title: "Data Preparation and Feature Engineering"
author: "Cameron Sterling"
date: "`r Sys.Date()`"
output: pdf_document
---
Here, you can find the code necessary to prepare the data and do some basic quality checks. 

```{r setup, include=FALSE}
library(tidyverse)
library(readr)
library(jsonlite)
library(caret)
library(randomForest)
library(tidycensus)
library(sf)
library(xgboost)
library(httr)
library(VIM)
library(FNN)
library(DMwR2)

```



# Loading the Data
## ACS Loading and Feature Engineering
```{r}
library(tidycensus)
library(tidyverse)

ny_metro_fips <- c(
  "36005", "36047", "36061", "36081", "36085", 
  "36059", "36103", "36119", "36087", 
  "34003", "34013", "34017", "34023", "34025", "34027", "34029", "34031", "34035", "34039" 
)


rent_burden_ny <- get_acs(
  geography = "tract",
  variables = c(
    total_renters = "B25070_001E",
    rent_burden_30_34 = "B25070_007E",
    rent_burden_35_39 = "B25070_008E", 
    rent_burden_40_49 = "B25070_009E", 
    rent_burden_50_plus = "B25070_010E"
  ),
  state = "36",
  county = c("005", "047", "061", "081", "085", 
  "059", "103", "119", "087"), 
  year = 2022,  
  survey = "acs5",
  output = "wide"
)

rent_burden_nj <- get_acs(
  geography = "tract",
  variables = c(
    total_renters = "B25070_001E",
    rent_burden_30_34 = "B25070_007E",
    rent_burden_35_39 = "B25070_008E", 
    rent_burden_40_49 = "B25070_009E", 
    rent_burden_50_plus = "B25070_010E"
  ),
  state = "34",
  county = c( "003", "013", "017", "023", "025", "027", "029", "34031", "035", "039"),  
  year = 2022,  
  survey = "acs5",
  output = "wide"
)

rent_burden_df <- bind_rows(rent_burden_ny, rent_burden_nj)

income_ny <- get_acs(
  geography = "tract",
  variables = c(median_income = "B19013_001E"),
  state = c("36"),
  county = c(
  "005", "047", "061", "081", "085", 
  "059", "103", "119", "087"
),
  year = 2022,
  survey = "acs5",
  output = "wide"
)

income_nj <- get_acs(
  geography = "tract",
  variables = c(median_income = "B19013_001E"),
  state = c("34"),
  county = c(
  "003", "013", "017", "023", "025", "027", "029", "031", "035", "039" 
),
  year = 2022,
  survey = "acs5",
  output = "wide"
)

income <- bind_rows(income_ny, income_nj)


tenure_ny <- get_acs(
  geography = "tract",
  variables = c(
    total_hh = "B25003_001E", 
    owner_hh = "B25003_002E",
    renter_hh = "B25003_003E"
  ),
  state = c("36"),
  county = c(
  "005", "047", "061", "081", "085", 
  "059", "103", "119", "087"
),
  year = 2022,
  survey = "acs5",
  output = "wide"
)


tenure_nj <- get_acs(
  geography = "tract",
  variables = c(
    total_hh = "B25003_001E", 
    owner_hh = "B25003_002E",
    renter_hh = "B25003_003E"
  ),
  state = c("34"),
  county = c(
  "003", "013", "017", "023", "025", "027", "029", "031", "035", "039" 
),
  year = 2022,
  survey = "acs5",
  output = "wide"
)
tenure <- bind_rows(tenure_ny, tenure_nj)

acs_data <- rent_burden_df %>%
  left_join(income, by = c("GEOID", "NAME")) %>%
  left_join(tenure, by = c("GEOID", "NAME"))

acs_data <- acs_data %>%
  mutate(

    rent_burden_30_plus = ifelse(renter_hh > 0, 
                                 (rent_burden_30_34 + rent_burden_35_39 + rent_burden_40_49 + rent_burden_50_plus) / renter_hh * 100, 
                                 NA),
    
    severe_rent_burden_50_plus = ifelse(renter_hh > 0, 
                                        rent_burden_50_plus / renter_hh * 100, 
                                        NA),
    
    percent_renters = ifelse(total_hh > 0, 
                             renter_hh / total_hh * 100, 
                             NA)
  ) %>%
  select(GEOID, NAME, median_income, percent_renters, rent_burden_30_plus, severe_rent_burden_50_plus)

acs_data <- acs_data %>%
  rename(
    percent_rent_burdened = rent_burden_30_plus,
    percent_severely_burdened = severe_rent_burden_50_plus
  )

```

## SVI Data
You can download this data on the CDC's website: https://www.atsdr.cdc.gov/place-health/php/svi/index.html#:~:text=Social%20vulnerability%20refers%20to%20the,and%20other%20community%2Dlevel%20stressors. 
```{r}
svi_ny <- read_csv("SVI_NJ_2.csv")
svi_nj <- read_csv("SVI_NY.csv")

svi_data <- bind_rows(svi_ny, svi_nj)
svi_selected <- svi_data %>%
  select(
    GEOID = FIPS,  
    total_population = E_TOTPOP,
    total_housing_units = E_HU,
    total_households = E_HH,
    percent_poverty_150 = EP_POV150,  
    percent_unemployed = EP_UNEMP,  
    percent_multiunit = EP_MUNIT,
    percent_mobile = EP_MOBILE,
    percent_no_hs_diploma = EP_NOHSDP,  
    percent_uninsured = EP_UNINSUR,  
    percent_seniors = EP_AGE65,  
    percent_youth = EP_AGE17,  
    percent_disability = EP_DISABL,  
    percent_single_parent = EP_SNGPNT,  
    percent_minority = EP_MINRTY,  
    percent_black = EP_AFAM,  
    percent_hispanic = EP_HISP,  
    percent_asian = EP_ASIAN,  
    percent_native = EP_AIAN,  
    percent_pacific_islander = EP_NHPI,  
    percent_two_or_more_races = EP_TWOMORE,  
    percent_limited_english = EP_LIMENG,  
    percent_large_apartments = EP_MUNIT,  
    percent_overcrowded = EP_CROWD,  
    percent_no_vehicle = EP_NOVEH,  
    percent_group_quarters = EP_GROUPQ,  
    percent_no_internet = EP_NOINT,  
    daytime_population_change = E_DAYPOP  
  )
```
```{r}
svi_selected$GEOID <- as.character(svi_selected$GEOID)

acs_data <- acs_data %>%
  left_join(svi_selected, by = "GEOID")


```
## EJI (CDC)
You can find this data on the CDC's website (for now): https://www.atsdr.cdc.gov/place-health/php/eji/eji-data-download.html
```{r}
eji_ny <- read.csv("EJI_2022_New_York.csv")
eji_nj <- read.csv("EJI_2022_New_Jersey.csv")
eji_data <- bind_rows(eji_ny, eji_nj)

eji_filtered <- eji_data %>%
  select(
    GEOID, 
    air_pollution_pm25 = E_PM,  
    air_pollution_ozone = E_OZONE,  
    diesel_exposure = E_DSLPM,  
    cancer_risk = EP_CANCER,  
    green_space = E_PARK,
    ozone_percentile = E_OZONE,  
    lead_exposure = E_HOUAGE,
    toxic_exposure = E_TRI,
    polluted_water = E_IMPWTR
  
  )
eji_filtered$GEOID <- as.character(eji_filtered$GEOID)
merged_data <- acs_data %>%
  left_join(eji_filtered, by = "GEOID")


```

# Basic Quality Checks
## Checking for NA Values and Removing Poorly Reported Rows 

```{r}
dim(merged_data)
sum(is.na(merged_data$percent_rent_burdened))
sum(is.na(merged_data$lead_exposure))
na_threshold <- ncol(merged_data) * 0.3
merged_data[merged_data == -999.0] <- NA

merged_clean <- merged_data[rowSums(is.na(merged_data)) <= na_threshold, ]

dim(merged_clean)

```

## Looking at the Distribution of the Outcome
```{r}
summary(merged_data$percent_rent_burdened)
summary(merged_data$percent_severely_burdened)


sum(merged_data$percent_rent_burdened < 0, na.rm = TRUE)

sum(merged_data$E_TOTPOP < 0, na.rm = TRUE)

ggplot(merged_data, aes(y = percent_severely_burdened)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(title = "Distribution of Rent Burden (%)", y = "Rent Burden (%)") +
  theme_minimal()
```

## Seeing Distribution of Outcome Geographically (Do I have any holes?)
```{r}
library(tigris)
merged_data_final <- numeric_data
heatmap_data <- merged_data_final %>%
  select(GEOID, percent_rent_burdened, percent_severely_burdened) %>%
  na.omit()
tracts_ny <- tracts(state = "NY", county = c("005", "047", "061", "081", "085", "059", "103", "119", "087"), year = 2022)
tracts_nj <- tracts(state = "NJ", county = c("003", "013", "017", "023", "025", "027", "029", "031", "035", "039"), year = 2022)

tracts <- rbind(tracts_ny, tracts_nj)
map_data <- left_join(tracts, heatmap_data, by = "GEOID")

ggplot(data = map_data) +
  geom_sf(aes(fill = percent_rent_burdened), color = NA) + 
  scale_fill_viridis_c(option = "plasma", name = "% Rent Burdened") 
  labs(
    title = "Rent Burden Across Census Tracts in NY & NJ",
    subtitle = "Percentage of Renters Paying >30% of Income on Rent",
  ) +
  theme_minimal()
  
ggplot(data = map_data) +
  geom_sf(aes(fill = percent_severely_burdened), color = NA) + 
  scale_fill_viridis_c(option = "plasma", name = "% Severly Rent Burdened") 
  labs(
    title = "Rent Burden Across Census Tracts in NY & NJ",
    subtitle = "Percentage of Renters Paying >50% of Income on Rent",
  ) +
  theme_minimal()
```


```
